---
title: "NOAA Storm Impacts Analysis"
author: "Javier Quintero Ramírez"
date: "30/12/2020"
output:  
    html_document:
    keep_md: true
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## Analysis of the impact of storms on public health and the economy in the United States from 1950 to 2011 

## ¬ Synopsis

#### The basic objective of this analysis is to explore the National Oceanic and Atmospheric Administration (NOAA) storm database and answer two basic questions about severe weather events:

##### 1. Across the United States, what types of weather events are most damaging to the health of the population?

##### 2. Across the United States, what types of weather events have the greatest economic consequences?

#### NOAA's database can be downloaded from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). It contains information about the severe weather events that occurred between 1950 and 2011 and their relationship to public health and economy of the United States in that time period.

#### For further information, here you can see the National Weather Service Storm Data [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) and the National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).

    
## ¬ Data Processing    
    
#### First, the NOOA's database is downloaded and read it:
```{r download_unzip_database, cache = TRUE}
if (!file.exists('./resources/repdata-data-StormData.csv.bz2')) 
{
   download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                  destfile = "./resources/repdata-data-StormData.csv.bz2")
}

stormDataBase <- read.csv("./resources/repdata-data-StormData.csv")

```

### Summary
#### Here we can see a summary and the structure of the data:
```{r sumary, cache = TRUE}
str(stormDataBase)
summary(stormDataBase)
```


#### Here are described the relevant variables for this analysis:

Health-related variables:

    FATALITIES: Number of deaths.
    INJURIES:   Number of injuries.

Economic-related variables:

    PROPDMG:    Property damaged.
    PROPDMGEXP: The units for property damage value.
    CROPDMG:    Crop damaged.
    CROPDMGEXP: The units for crop damage value.

And these is the number of weather events types found in the NOOA's database:
    
```{r evtype1}    
evtypeList <- unique(toupper(stormDataBase$EVTYPE))
length(evtypeList)
```

#### Now we are going to select only the relevants variables:
```{r select_Data}    
selectedVariables <- select(stormDataBase,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,EVTYPE)
head(selectedVariables)
```

#### As there are many types of climatic events, we will try to unify some of them, which have a certain relationship:
```{r filter_Type}

selectedVariables$EVTYPE <- gsub('.*SUMMARY*', 'SUMMARY',toupper(selectedVariables$EVTYPE))

selectedVariables$EVTYPE <- gsub('.*DRY.*', 'DROUGHT', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*DROUGHT.*', 'DROUGHT', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*DRIE.*', 'DROUGHT', toupper(selectedVariables$EVTYPE))

selectedVariables$EVTYPE <- gsub('.*COLD.*', 'FROST', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*LOW.*TEMPER.*', 'FROST', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*FROST.*', 'FROST', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*FREEZ.*', 'FROST', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*HAIL.*', 'FROST', toupper(selectedVariables$EVTYPE))

selectedVariables$EVTYPE <- gsub('.*HIGH.*TEMPER.*', 'HEAT WAVES', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*HEAT.*', 'HEAT WAVES', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*WARM.*', 'HEAT WAVES', toupper(selectedVariables$EVTYPE))

selectedVariables$EVTYPE <- gsub('.*HURRICANE.*', 'HURRICANE', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*FIRE.*', 'FIRE', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*FLOOD.*', 'FLOOD', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*TORN.*', 'TORNADO', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*WIND.*', 'WIND', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*RAIN.*', 'RAIN', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*STORM.*', 'STORM', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- gsub('.*SNOW.*', 'SNOW', toupper(selectedVariables$EVTYPE))
selectedVariables$EVTYPE <- trimws(selectedVariables$EVTYPE)

filteredData <- filter(selectedVariables, EVTYPE != "SUMMARY" && EVTYPE != "?")
remove(selectedVariables)
```

```{r evtype12}
evtypeList2 <- unique(filteredData$EVTYPE)
l <- length(evtypeList2)
```
This leaves us as a result `r l` different types of events

#### Now, since there are different types of units of value for recording damage, we go on to create new variables to standardize the value units in dollars:
```{r unify_value, warning=FALSE}
filteredData$VALUEPROPDMG <- filteredData$PROPDMG
filteredData$VALUEPROPDMG[which(toupper(filteredData$PROPDMGEXP) == "K")] <- filteredData$PROPDMG * 1000 
filteredData$VALUEPROPDMG[which(toupper(filteredData$PROPDMGEXP) == "M")] <- filteredData$PROPDMG * 1000000
filteredData$VALUEPROPDMG[which(toupper(filteredData$PROPDMGEXP) == "B")] <- filteredData$PROPDMG * 1000000000

filteredData$VALUECROPDMG <- filteredData$CROPDMG
filteredData$VALUECROPDMG[which(toupper(filteredData$CROPDMGEXP) == "K")] <- filteredData$CROPDMG * 1000
filteredData$VALUECROPDMG[which(toupper(filteredData$CROPDMGEXP) == "M")] <- filteredData$CROPDMG * 1000000
filteredData$VALUECROPDMG[which(toupper(filteredData$CROPDMGEXP) == "B")] <- filteredData$CROPDMG * 1000000000

filteredData$VALUE = filteredData$PROPDMG + filteredData$CROPDMG

unifiedData <- select(filteredData,FATALITIES,INJURIES,VALUE, EVTYPE)
remove(filteredData)
head(unifiedData)
```

## ¬ Results
### Question 1: 
#### Across the United States, what types of weather events are most damaging to the health of the population?
##### We have two measures of health effects: deaths and injuries
##### Here we can see the response related to Deaths
```{r questionDeath, message=FALSE}
library(ggplot2)

unifiedData %>%
      select(FATALITIES, EVTYPE) %>%
      group_by(EVTYPE) %>%
      summarise(sumF = sum(FATALITIES)) %>%
      top_n( n = 5, wt = sumF) %>%
      ggplot(aes(y = sumF, x = reorder(x = EVTYPE, X = sumF), fill = EVTYPE)) +
      geom_bar(stat = "identity") +
      xlab(label = "Top 5 types of most damaging weather events") +
      ylab(label = "Number of Deaths") 
```

##### And here we can see the response related to Injuries
```{r questionInjuries, message=FALSE}
unifiedData %>%
      select(INJURIES, EVTYPE) %>%
      group_by(EVTYPE) %>%
      summarise(SumI = sum(INJURIES)) %>%
      top_n(n = 5, wt = SumI) %>%
      ggplot(aes(y = SumI, x = reorder(x = EVTYPE, X = SumI), fill = EVTYPE))+
      geom_bar(stat = "identity") +
      xlab(label = "Top 5 types of most damaging weather events") +
      ylab(label = "Number of Injuries Cases")
```

### Question 2: 
#### Across the United States, what types of weather events have the greatest economic consequences?
##### In the following plot we can see which climatic events had the most impact on the United States economy in the period studied:
```{r questionEconomic}
unifiedData %>%
      select(VALUE, EVTYPE) %>%
      group_by(EVTYPE) %>%
      summarise(SumV = sum(VALUE)/1000000) %>%
      top_n(n = 5, wt = SumV) %>%
      ggplot(aes(y = SumV, x = reorder(x = EVTYPE, X = SumV), fill = EVTYPE)) +
      scale_color_discrete(name="Weather Events") + 
      geom_bar(stat = "identity") +
      xlab(label = "Top 5 types of most expensive weather events") +
      ylab(label = "Millons of Dollars") 
```